A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-11-08 analysis/paper/paper.Rmd
Ragnhildrød will take some time

# Introduction
The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout history, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method has long history of use in the region, as is frequently applied within both the research and heritage management sector to date diverse archaeological phenomena such as rock art, harbour constructions and, as is the focus of this study, Stone Age sites,

@@

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.  If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify societal shifts, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty through Monte Carlo simulation, a similar approach is adopted here.

# Background
Relative sea level is so named because the parameters of interest is the elevation of the sea surface relative to land. Variation in this relative distance follows from effects that can be divided into three main categories. The first of these is eustatic sea-level change. Eustatic sea-level is understood as the sea level if the water has been evenly distributed across the Earth's surface and without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust which causes its weight to be redistributed. These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

@@

Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory, a process that also continues to this day. In other areas of Norway the situation is sometimes more complicated. On the outer coast, the relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed. In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle. Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span. As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.

The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this. In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was *not* shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the ledge and the site would have been accessible by means of watercraft. This line of reasoning is often also used in survey projects which typically do not produce material suitable for radiocarbon dating. The evaluation of the degree to which sites were shorebound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and collegues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found a overlap of the confidence bands of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites. However, in several of the cases where there was a discrepancy, the

@@

This study necessaitated the compilation of multiple relative sea-level curves developed for different parts of the region. Most of the curves come provided with a confidence interval which is largely based on an evaluation by the geologists who developed the curve. The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples,  direction and adjustment to a common isobase based on the direction and tilt of the variation in rebound intensity, as well as, to name but a few. The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods, and often using older and outdated calibration curves for both marine and terrestrial samples. However, any reinterpretation and standardisation of the curves is also hampered by a range of issues. The original data or necessary details about the sampling processes are not always available, and a reinterpretation based soley on radiometric dates, without taking into account the considerations and evaluations of the field geologists can be entirely misleading. In a rudimentary attempt to account for these differences when compiling the curves, the width of the confidence band for the sea level at a single date across the entire region is given as the largerst uncertainty of any one of the curves at that point in time. This averages to an uncertainty of +/- calendar years. This uncertainty is uniformly distributed within the confidence band.

# Data
All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive

The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west. These number a total nrow() sites, of which nrow() have radiocarbon dates. Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below). Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the MCH. Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.

@@

For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve using OxCal v. . All dates not younger than 1700 BCE with 98 % certainty were then excluded from further analysis. Radiocarbon dates associated with each site were then inspected, and with reference to the evaluation done in the original reports, dates believed to belong to the same settlement phase were grouped. In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal. Multiple phases at a single site were treated as independent of each other (Figure x).

@@

Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern, the simulation was performed. Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase. This year then has a corresponding likely elevation range of the shoreline. An elevation value was drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing, and creating polygons of the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDist package for R. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.

>>>>

# Introduction
The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method has long history of use in the region, as is frequently applied to assign an approximate date to diverse archaeological phenomena such as rock art, harbour and other sea-side constructions and, as is the focus of this study, Stone Age sites.

@@

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty through Monte Carlo simulation, a similar approach is adopted here. The results of this study can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.  If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify important societal changes, and any such variation can potentially also have implications for how archaeological investigations should be directed within the area of analysis.

# Background
Relative sea-level is so named because the parameters of interest is the elevation of the sea surface relative to land. Variation in this relative distance follows from effects that can be divided into three main categories. The first of these is changes to the eustatic sea-level, which is understood as the sea-level if the water has been evenly distributed across the Earth's surface without adjusting for variation in the rigidity of the Earth, its rotation, or the self-gravitation inherent to the water body itself. The eustatic sea-level is mainly impacted by glaciation and de-glaciation, which can either bind or release large amounts of water into the oceans, as well as by volumetric changes in the Earth's crust following from . The second category is that of istostasy, which pertains to adjustments in the Earth's crust to regain gravitational equilibrium relative to the underlying viscous mantle. This is often the result of glacial istostasy which follows from glaciation and de-glaciation and corresponding loading and unloading of weight, as well as from erosion of the crust which causes its weight to be redistributed. These effects thus causes the lithosphere to be either pushed into the mantle due to increased weight, or to rebound and to lift upwards due to lower weight. Thirdly, relative sea-level is also impacted by thermostasy, which pertains to water temperatures and corresponding variation in water density.

@@

Following the end of the Weichselian and the retreat of the Fennoscandian Ice Sheet, the isostatic rebound in south-eastern Norway has been so severe that the region has been subject to a continuous sea-level regression, despite corresponding eustatic sea-level rise. In other words, the relative sea level has been dropping throughout prehistory---a process that also continues to this day. In other areas of Norway the situation is sometimes more complicated. On the outer coast, the relative sea-level change has transitioned between regression and transgression phases, and the archaeological remains can be destroyed. In other areas, the sea-level has been stable over longer periods of time, meaning coastal locations have often been reused multiple times over long time-spans, creating a palmpiset of settlement events that are impossible to disentangle. Comparatively, given the continuous and relatively rapid shoreline regression in south-eastern Norway, any one location in the region has only been shore-bound within a relatively limited time-span. As this general displacement trajectory is the same throughout the entirety of prehistory, this makes the region especially useful for testing of the assumption of a shore-bound settlement and to identify potential deviations from this pattern.

The most common way to evaluate whether or not a site would have been shorebound is typically through an evaluation of typological indicators in the assemblages and its correspondence with available shoreline displacement curves. If this corresponds then it is common to err on the side of a shorebound site location, even if the time-span of the dates provided by the typological indicators are too wide to decisively verify this. In the case that radiocarbon dates of site features are available, this is typically evaluated by means of a visual evaluation of the overlap between shoreline displacement curves and the radiocarbon probability distributions. Another line of reasoning has been to combine these approaches with a qualitative consideration of the landscape surrounding the sites, and an evaluation of the degree to which the site location would have been sensible if the site was *not* shore bound [e.g. @jaksland2014]. This can for example pertain to accessibility. If the site is situated on a ledge in a steep and jagged area of the present day landscape, it would make intuitive sense that the location was used when the ocean reached the elevation of the site which would then have been accessible by means of watercraft. This line of reasoning is often also used in survey projects, which typically do not produce material suitable for radiocarbon dating. The evaluation of the degree to which sites were shore-bound is therefore typically conducted on a site-by-site or project-by-project basis, meaning the fundamental basis for assuming shore bound site location is largely based on a mosaic of smaller scale investigations. Although it appears that the arguments for such site locations is sensible and for a large part appear to be correct, few attempts at a comprehensive evaluation and quantification of this tendency has been done.

One of the only larger scale evaulations of this relationship in Norwegian archaeology has been done by Solheim and collegues (@solheim20, see also @solheim2020), who compared 102 radiocarbon dates from 29 Mesolithic sites on the western side of the Oslo fjord to the displacement curve for the Larvik area. They found a overlap between the probability distribution of the radiocarbon dates with the shoreline displacement curve for 86.5 % of the sites. However, in several of the cases where there was a discrepancy, the

@@

This study necessaitated the compilation of multiple relative sea-level curves developed for different parts of the region. Most of the curves come provided with a confidence interval which is largely based on an evaluation by the geologists who developed the curve. The reason why these confidence intervals are not strictly mathematically derived is that the curves do not only contain uncertainy as related to radiometric dates, which are fairly well understood, but also hold potential error as related to the interpretation of sediment cores, placement of samples, as well as direction and adjustment to a common isobase based on the direction and gradient in rebound intensity, to name but a few. The error bands are consequently complex, and cannot necessarily be directly compared between curves as these have been developed at different times, using a wide range of different methods. However, any reinterpretation and standardisation of the curves is also hampered by a range of issues. The original data or necessary details about the sampling processes are not always available, and a reinterpretation based soley on radiometric dates, without taking into account the considerations and evaluations of the field geologists can be entirely misleading. In a rudimentary attempt to account for these differences when compiling the curves, the width of the confidence band for the sea level at a single date across the entire region is given as the largerst uncertainty of any one of the curves at that point in time. This averages to an uncertainty of +/- calendar years. This uncertainty is uniformly distributed within the confidence band.

# Data
All data that are required to run the analyses and the derived data are freely available in an online repository. The dataset is slightly more extensive, including excavated sites without radiocarbon dates, and radiocarbon dates to periods other than the Stone Age.

The data compiled for the analysis consists of all excavated Stone Age sites from the region between Horten county in the north-east, to Arendal in the south-west. These number a total nrow() sites, of which nrow() have radiocarbon dates. Of these, in turn, nrow() sites have radiocarbon dates to the Stone Age, and nrow() sites have radiocarbon dates to the Stone Age that were evaluated by the original excavators to relate to the lithic inventory on the sites (see also below). Spatial data in the form of features and site limits as defined by the excavation was retrieved from local databases at the Museum of Cultural History. Each radiocarbon date has been associated with site features or, where these weren't available, the spatial limit of the entire site. Due to somewhat variable practices between excavations, what spatial geometry best represents the site limit was decided based on an evaluation of the excavation reports. This means that the limits are variably given as area de-turfed before excavation, area stripped with excavator following the excavation, manually excavated area, or convex hulls generated around the site features.

@@

For the date ranges associated with the sites, all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve using OxCal v. . All dates not younger than 1700 BCE with 99.7 % uncertainty range were then excluded from further analysis. Radiocarbon dates associated with each site were then inspected, and with reference to the evaluation done in the original reports, dates believed to belong to the same settlement phase were grouped. In the case were there are multiple dates believed to belong to a single phase, these were subjected to Bayesian modelling using the Boundary function in OxCal. Multiple phases at a single site were treated as independent of each other (Figure x).

@@

Armed with a likely date range for the occupation(s) of each site, the likely trajectory of relative sea-level change at that location, and an DTM edited to remove substantial modern disturbances, the simulation was performed. Each simulation run involved first drawing a single year from the posterior density estimate of a given occupation phase. This year then has a corresponding likely elevation range of the shoreline. An elevation value was drawn uniformly from this range, using intervals of 5 cm. The sea-level was then raised to this elevation on the DTM by defining all elevation values at or below this altitude as missing. Polygons were then created from the resulting areas with missing values. The horisontal distance was then found by measuring the shortest distance between site and sea polygons, and the vertical distance by subtracting the elevation of the sea-level from the lowest elevation of the site polygon. The so-called topographic distance between site and sea was also found by measuring the distance while taking into account the slope of the terrain on the DTM. This was done using the topoDist package for R. In the case where the sea-polygons intersects the site polygon, all distance measures were set to zero. In the case that the sea-polygons completely contain the site, the horisontal and topographic distance measures were made negative. While its safe to assume that an archaeological site was not occupied when it was located beneath sea-level, a negative result can reflect the inherent uncertainty in this procedure and might also help identify discrepancies in displacement data or radiocarbon dates.
